Sparse spectral and $p$-finite element methods for partial differential equations on disk-slices and trapeziums

### 2019-04-15 papers/halfdisk/halfdisk.tex
New figures. Comments added.

The weight $w_2(x)$ is an ultraspherical weight, and the corresponding OPs are   the Jacobi polynomials  \(\{P_n^{(b, b)}\}\). The weight $w_1(x)$ is non-classical (it is in fact semi-classical, and is equivalent to a generalized Jacobi weight \cite[\S5]{magnus1995painleve}) and we introduce the notation \(\{\genjac_n^{(a, b)}\}\) for the orthonormal polynomials. Thus we arrive at  a two parameter family of 2D orthogonal polynomials\sotodo{We should probably use a different letter than $P$ to avoid confusion with Jacobi polynomials}
\ \(\{\hdopnkab\}\) given by, for \(0 \le k \le n, \: n = 0,1,2,\dots,\)
\begin{align}\label{eq:diskpolys}
 \hdopnk^{(a,b)}(x,y) := \genjacnmk^{(a, b+k+\half)}(x) \: \rho(x)^k \: P_k^{(b,b)}\fpr(\frac{y}{\rho(x)}), \quad (x,y) \in \Omega, 
\end{align}
orthogonal on \(\Omega\) with respect to the weight
\begin{align}
W^{(a,b)}(x,y) = \genjacw^{(a,b)}(x) \: w_P^{(b)}(\frac{y}{\rho(x)}) = x^a \: (1-x^2-y^2)^b.
\end{align}

@@

We can express the three-term recurrences associated to $\genjac_n^{(a,b)}$ and $P_n^{(b,b)}$ as
\begin{align}
x \genjac_n^{(a,b)}(x) &= \beta_n^{(a,b)} \genjac_{n+1}^{(a,b)}(x) + \alpha_n^{(a,b)} \genjac_n^{(a,b)}(x) + \beta_{n-1}^{(a,b)} \genjac_{n-1}^{(a,b)}(x) 
\label{eqn:Hrecurrence} \\
y P_n^{(b,b)}(y) &= \delta_n^{(b)} P_{n+1}^{(b,b)}(y) + \gamma_n^{(b)} P_n^{(b,b)}(y) + \delta_{n-1}^{(b)} P_{n-1}^{(b,b)}(y),
\end{align}
where we note that \(\gamma_n^{(b)} = 0\), for \( n = 0,1,2,\dots\). We can use these to determine the 2D recurences for $\hdopnkab(x,y)$:

@@

\begin{figure}
	\begin{subfigure}[t]{0.3\textwidth}
	\centering
	\includegraphics[scale=0.3]{solution-poisson}
	\caption{}
	\label{fig:solution-poisson}
	\end{subfigure}
	\begin{subfigure}[t]{0.3\textwidth}
	\centering
	\includegraphics[scale=0.3]{solution-helmholtz}
	\caption{}
	\label{fig:solution-helmholtz}
	\end{subfigure}
	\begin{subfigure}[t]{0.3\textwidth}
	\centering
	\includegraphics[scale=0.3]{solution-biharmonic}
	\caption{}
	\label{fig:solution-biharmonic}
	\end{subfigure}
	\centering
    	\caption{(a): The computed solution to $\Delta u = f$ with zero boundary conditions with $f(x,y) = 1 + \text{erf}(5(1 - 10((x - 0.5)^2 + y^2)))$. (b): The computed solution to $\Delta u + k^2 \: v \: u = f$ with zero boundary conditions with $f(x,y) = xye^x$, $v(x,y) = xy^2$ and $k = 0.25$ \sotodoinline{Make a lot higher}. (c): The computed solution to $\Delta^2 u = f$ with zero Dirichlet and Neumann boundary conditions with $f(x,y) = 1 + \text{erf}(5(1 - 10((x - 0.5)^2 + y^2)))$.}
        \label{fig:solution}
\end{figure}

@@

We can tackle the problem as follows. Fix \(a, b = 1,1\). Denote the coefficient vector for expansion of $u$ in the $\bigWii$ OP basis up to degree $N$ by $\mathbf{u}$, and the coefficient vector for expansion of $f$ in the $\bighdopii$ OP basis up to degree $N$ by $\mathbf{f}$. Since $f$ is known, we can obtain $\mathbf{f}$ using the quadrature rule above. In matrix-vector notation, our system hence becomes:
\begin{align}
    \laplacewii \mathbf{u} = \mathbf{f}
\end{align}
which can be solved to find $\mathbf{u}$.
In Figure \ref{fig:solution-poisson} we see the solution to the Poisson equation with zero boundary conditions given in (\ref{eqn:poisson}) in the half-disk $\Omega$. In Figure \ref{fig:solutionblocknorms} we show the norms of each block of calculated coefficients of the approximation for four right-hand sides of the Poisson equation with N = 200, that is, 20,301 unknowns. The rate of decay in the coefficients is a proxy for the rate of convergence of the computed solution. We see that\bstodo{Describe what plot tells us}

\begin{figure}
        	\includegraphics[scale=0.5]{solutionblocknorms}
	\centering
	\caption{The norms of each block of the computed solution of the Poisson equation with the given right hand side functions. This demonstrates algebraic convergence with the rate dictated by the decay at the corners, with spectral convergence observed when the right-hand side vanishes to all orders.  \sotodoinline{More blocks}}
        	\label{fig:solutionblocknorms}
\end{figure}

>>>>

The weight $w_2(x)$ is an ultraspherical weight, and the corresponding OPs are   the Jacobi polynomials  \(\{P_n^{(b, b)}\}\). The weight $w_1(x)$ is non-classical (it is in fact semi-classical, and is equivalent to a generalized Jacobi weight \cite[\S5]{magnus1995painleve}) and we introduce the notation \(\{\genjac_n^{(a, b)}\}\) for the orthonormal polynomials. Thus we arrive at  a two parameter family of 2D orthogonal polynomials $\{\hdopnkab\}$ given by, for $0 \le k \le n, \: n = 0,1,2,\dots,$
\begin{align}\label{eq:diskpolys}
 \hdopnk^{(a,b)}(x,y) := \genjacnmk^{(a, b+k+\half)}(x) \: \rho(x)^k \: P_k^{(b,b)}\fpr(\frac{y}{\rho(x)}), \quad (x,y) \in \Omega, 
\end{align}
orthogonal on \(\Omega\) with respect to the weight
\begin{align}
W^{(a,b)}(x,y) = \genjacw^{(a,b)}(x) \: w_P^{(b)}(\frac{y}{\rho(x)}) = x^a \: (1-x^2-y^2)^b.
\end{align}

@@

We can express the three-term recurrences associated to $\genjac_n^{(a,b)}$ and $P_n^{(b,b)}$ as
\begin{align}
x \genjac_n^{(a,b)}(x) &= \beta_n^{(a,b)} \genjac_{n+1}^{(a,b)}(x) + \alpha_n^{(a,b)} \genjac_n^{(a,b)}(x) + \beta_{n-1}^{(a,b)} \genjac_{n-1}^{(a,b)}(x) 
\label{eqn:Hrecurrence} \\
y P_n^{(b,b)}(y) &= \delta_n^{(b)} P_{n+1}^{(b,b)}(y) + \gamma_n^{(b)} P_n^{(b,b)}(y) + \delta_{n-1}^{(b)} P_{n-1}^{(b,b)}(y),
\end{align}
where we note that \(\gamma_n^{(b)} = 0\), for \( n = 0,1,2,\dots\). We can use these to determine the 2D recurrences for $\hdopnkab(x,y)$:

@@

\begin{figure}
	\begin{subfigure}[t]{0.3\textwidth}
	\centering
	\includegraphics[scale=0.3]{solution-poisson}
	\caption{}
	\label{fig:solution-poisson}
	\end{subfigure}
	\begin{subfigure}[t]{0.3\textwidth}
	\centering
	\includegraphics[scale=0.3]{solution-helmholtz}
	\caption{}
	\label{fig:solution-helmholtz}
	\end{subfigure}
	\begin{subfigure}[t]{0.3\textwidth}
	\centering
	\includegraphics[scale=0.3]{solution-biharmonic}
	\caption{}
	\label{fig:solution-biharmonic}
	\end{subfigure}
	\centering
    	\caption{(a): The computed solution to $\Delta u = f$ with zero boundary conditions with $f(x,y) = 1 + \text{erf}(5(1 - 10((x - 0.5)^2 + y^2)))$. (b): The computed solution to $\Delta u + k^2 \: v \: u = f$ with zero boundary conditions with $f(x,y) = xye^x$, $v(x,y) = xy^2$ and $k = 200$. (c): The computed solution to $\Delta^2 u = f$ with zero Dirichlet and Neumann boundary conditions with $f(x,y) = 1 + \text{erf}(5(1 - 10((x - 0.5)^2 + y^2)))$.}
        \label{fig:solution}
\end{figure}

@@

We can tackle the problem as follows. Fix \(a, b = 1,1\). Denote the coefficient vector for expansion of $u$ in the $\bigWii$ OP basis up to degree $N$ by $\mathbf{u}$, and the coefficient vector for expansion of $f$ in the $\bighdopii$ OP basis up to degree $N$ by $\mathbf{f}$. Since $f$ is known, we can obtain $\mathbf{f}$ using the quadrature rule above. In matrix-vector notation, our system hence becomes:
\begin{align}
    \laplacewii \mathbf{u} = \mathbf{f}
\end{align}
which can be solved to find $\mathbf{u}$.
In Figure \ref{fig:solution-poisson} we see the solution to the Poisson equation with zero boundary conditions given in (\ref{eqn:poisson}) in the half-disk $\Omega$. In Figure \ref{fig:solutionblocknorms} we show the norms of each block of calculated coefficients of the approximation for four right-hand sides of the Poisson equation with N = 200, that is, 20,301 unknowns. The rate of decay in the coefficients is a proxy for the rate of convergence of the computed solution. We see that... \bstodoinline{Describe what plot tells us - I dont want to copy and paste your Triangle paper here but not sure what else to write}

\begin{figure}
        	\includegraphics[scale=0.5]{solutionblocknorms}
	\centering
	\caption{The norms of each block of the computed solution of the Poisson equation with the given right hand side functions. This demonstrates algebraic convergence with the rate dictated by the decay at the corners, with spectral convergence observed when the right-hand side vanishes to all orders.  \bstodoinline{N=200 blocks here. When I have the time to run N=1000 or so I will do}}
        	\label{fig:solutionblocknorms}
\end{figure}
